Predicting Political Support in the United States

### 2024-03-16 output/paper.qmd
Format sections and complete results

# Data

@@

@@

    -   This data was represented in the form of a numerical variable corresponding to the gender of the voter. This was cleaned to limit the options to "Man", "Woman", and "Other".

@@

    -   This data was represented in the form of a numerical variable corresponding to the ethnicity of the voter. This was cleaned to limit the options to "White", "Black", "Hispanic", "Asian", "Middle Eastern", "Native American", "Multiracial", and "Other".

@@

    -   This data was represented in the form of a numerical variable corresponding to the religion of the voter. This was cleaned to limit the options to "Protestant", "Roman Catholic", "Jewish", "Atheist", "Agnostic", "Non-Religious", and "Other". There were various other religions that were not included in the dataset, but represented a small portion of the sample, and were thus grouped into "Other".

@@

@@

    Incorporating the voter's `religion`, was a needed factor since it is important to understand that religious beliefs and differences can have a strong influence one's personal beliefs and therefore impact their personal stances on topics such as gun ownership. To put a contrasting example out there, 30% of Protestant/Christian respondents support the Democrats as opposed to 72% by Atheists. This could be attributed to certain parties aligning more to certain religious beliefs. To see a bigger insight into this, take a look at @fig-racevsreligion1 and @fig-racevsreligion2 (figures separated according to respondent count), and notice examples like, 'White Protestants' appear significantly more supportive towards the Republicans as opposed to 'White Catholics' and 'Black Protestants' who appear to be significantly more supportive towards the Democrats.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Votes for Candidate given Race and Religion"
#| label: fig-racevsreligion1

ces2022_2 |>
  filter(race %in% c("Black", "Hispanic", "White")) |>
  ggplot(aes(x = religion, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(race)) +
  theme_minimal() +
  labs(
    x = "Religion",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Biden" = "blue", "Trump" = "red")) +
  theme(legend.position = "bottom", axis.text.x=element_text(angle=60, hjust=1))

@@

@@

@@

# Rename coefficients
cm <- c('genderMale' = 'Male',
        'genderOther' = 'Other Gender',
        'raceOther' = 'Other Race',
        'raceWhite' = 'White',
        'raceNative American' = 'Native American', 
        'raceMultiracial' = 'Multiracial',
        'raceMiddle Eastern' = 'Middle Eastern',
        'raceHispanic' = 'Hispanic',
        'raceBlack' = 'Black',
        'raceAsian' = 'Asian',
        'religionNone' = 'Non-Religious',
        'religionProtestant' = 'Protestant',
        'religionAgnostic' = 'Agnostic',
        'religionAtheist' = 'Atheist',
        'religionRoman Catholic' = 'Roman Catholic',
        'religionJewish' = 'Jewish',
        'religionOther' = 'Other',
        'age_group30-44' = 'Age (30-44)',
        'age_group45-65' = 'Age (45-65)',
        'age_group65+' = 'Age (65+)'
        )

@@

@@

@tbl-modelsummarycultural

@@

@@

@@

@@

cm <- c('genderMale' = 'Male',
        'raceWhite' = 'White',
        'raceNative American' = 'Native American', 
        'raceMultiracial' = 'Multiracial',
        'raceMiddle Eastern' = 'Middle Eastern',
        'raceHispanic' = 'Hispanic',
        'raceBlack' = 'Black',
        'raceAsian' = 'Asian',
        'religionNone' = 'Non-Religious',
        'religionProtestant' = 'Protestant',
        'religionAgnostic' = 'Agnostic',
        'religionAtheist' = 'Atheist',
        'religionRoman Catholic' = 'Roman Catholic',
        'religionJewish' = 'Jewish',
        'religionOther' = 'Other',
        'age_group30-44' = 'Age (30-44)',
        'age_group45-65' = 'Age (45-65)',
        'age_group65+' = 'Age (65+)'
        )

>>>>

# Data {#data}

@@

## Variables

@@

    -   This data was represented in the form of a numerical variable corresponding to the gender of the voter. This was cleaned to limit the options to "Man", "Woman", and "Other". Other includes both respondents who specifically selected "Other" as well as respondents who responded as "Non-binary". Non-binary responses only totaled 354 participants out of 60,000.

@@

    -   This data was represented in the form of a numerical variable corresponding to the ethnicity of the voter. This was cleaned to limit the options to "White", "Black", "Hispanic", "Asian", "Middle Eastern", "Native American", "Multiracial" and "Other".

@@

    -   This data was represented in the form of a numerical variable corresponding to the religion of the voter. This was cleaned to limit the options to "Protestant", "Roman Catholic", "Jewish", "Atheist", "Agnostic", "Non-Religious", and "Other". There were various other religions that were included in the dataset, but represented a small portion of the sample, and were thus grouped into "Other".

@@

## Justification

@@

    Incorporating the voter's `religion`, was a needed factor since it is important to understand that religious beliefs and differences can have a strong influence one's personal beliefs and therefore impact their personal stances on topics such as gun ownership. To put a contrasting example out there, 30% of Protestant/Christian respondents support the Democrats as opposed to 72% by Atheists. This could be attributed to certain parties aligning more to certain religious beliefs. To see a bigger insight into this, take a look at @fig-racevsreligionlarge and @fig-racevsreligion2 (figures separated according to respondent count), and notice examples like, 'White Protestants' appear significantly more supportive towards the Republicans as opposed to 'White Catholics' and 'Black Protestants' who appear to be significantly more supportive towards the Democrats.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Votes for Candidate given Race and Religion"
#| label: fig-racevsreligionlarge
#| layout-ncol: 2

ces2022_2 |>
  filter(race %in% c("Black", "Hispanic")) |>
  ggplot(aes(x = religion, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(race)) +
  theme_minimal() +
  labs(
    x = "Religion",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Biden" = "blue", "Trump" = "red")) +
  theme(legend.position = "bottom", axis.text.x=element_text(angle=60, hjust=1))

ces2022_2 |>
  filter(race %in% c("White")) |>
  ggplot(aes(x = religion, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(race)) +
  theme_minimal() +
  labs(
    x = "Religion",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Biden" = "blue", "Trump" = "red")) +
  theme(legend.position = "bottom", axis.text.x=element_text(angle=60, hjust=1))

@@

political_preferences_race <-
  readRDS(file = "models/political_preferences_race.rds")

@@

## Cultural Model

@@

# Rename coefficients
cm <- c('genderMale' = 'Male',
        'genderOther' = 'Other Gender',
        'raceOther' = 'Other Race',
        'raceWhite' = 'White',
        'raceNative American' = 'Native American', 
        'raceMultiracial' = 'Multiracial',
        'raceMiddle Eastern' = 'Middle Eastern',
        'raceHispanic' = 'Hispanic',
        'raceBlack' = 'Black',
        'raceAsian' = 'Asian',
        'religionNone' = 'Non-Religious',
        'religionProtestant' = 'Protestant',
        'religionAgnostic' = 'Agnostic',
        'religionAtheist' = 'Atheist',
        'religionRoman Catholic' = 'Roman Catholic',
        'religionJewish' = 'Jewish',
        'religionOther' = 'Other',
        'age_group30-44' = 'Age (30-44)',
        'age_group45-65' = 'Age (45-64)',
        'age_group65+' = 'Age (65+)'
        )

@@

## Political Importances Model

@@

## Cultural Model

  @tbl-modelsummarycultural first suggests a that males were less likely to vote for Biden, and more likely to vote for Trump. Looking at @fig-agevsgender-1 this is not an obvious correlation to make with the votes for Biden and Trump being close for each age group, with Trump only winning more votes with the male and above 65 demographic. However, looking at the 90% confidence intervals in @fig-conf_cultural for the coefficients of that model, specifically that for being male, we can see that the model is 90% confident that this relationship is around strictly negative against voting for Biden. We also see the disagreement within the 45+ community that @fig-agevsgender-1 shows, and the agreement within the younger community (18 - 45), with @tbl-modelsummarycultural showing a large correlation between being 30-44 to supporting Biden, with being aged 45-65 or 65+ having very little correlation to support Biden, supported by the 90% confidence interval shown in @fig-conf_cultural for those coefficients spanning both negative and positive, suggesting an uncertain preference. Although the model included being: Female, Asian, 18-29, and Agnostic, in the intercept, for @fig-agevsgender-1, @fig-racevsreligion2, and @fig-racevsreligionlarge and @fig-racevsreligion2 respectively, it reasonable to say that each graph suggests that being Female, Asian, 18-29, and Agnostic has a correlation with voting for Biden, each demographic having a significant amount more votes for Biden than Trump.

  Now looking at the coefficients corresponding to race and religion, @tbl-modelsummarycultural suggests that respondents identifying themselves as being Black or Atheist were more likely to have voted for Biden. Additionally, as mentioned before those who are Agnostic or Asian as per @tbl-modelsummaryrace, who were included as the intercept are suggested to have voted for Biden. The respondents identifying as White, Hispanic, Middle Eastern, Multiracial, and Native American are suggested to be less likely to vote for Biden.

  Overall, this model reinforces our expectations that suggest that the cultural background of a person has a considerable effect on their expected vote. Religion and Race especially have considerable leanings to one candidate over the other, suggesting a strong impact of race and religion in political ideals.

## Political Importances Model

  An strong finding that @tbl-modelsummaryimportances displays is household gun ownership's effect on their vote. As mentioned in the [model](#Model)'s justification, Trump held a pro-gun-ownership stance, whilst Biden pushed for reform to gun laws. @tbl-modelsummaryimportances mirrors these stances, predicting that respondents who own or have people in their household own guns are much more likely to vote for Trump, who aligns with their own personal gun ownership stance. This is strongly supported by @fig-confimportances, which shows that the model is 90% confident that the likelihood of a respondent whose household owns guns is between around -0.9 to -1.8, a strongly negative correlation. This is supported by @fig-gunownvseducation showing that that most respondents who do not own guns, had voted for Biden, and Trump was only able to secure a majority vote for any level of education, when the respondent's household owned guns.

  Similarly, @tbl-modelsummaryimportances shows a correlation between having student loans and supporting Biden. Also mentioned in the [model](#Model)'s justification, the student loan forgiveness stance that more widely benefited current student loan holders was held by Biden. This fact is supported by the model that suggests a respondent who has student loans is more likely to support Biden over Trump. Being such an significant difference in everyday finances, whether they have to think about paying off student loans or not, it is reasonable to expect the outcome that respondents would want to vote for the candidate who can help take that burden off.

  It is also important to note education's impact on respondent's voting habits. The 90% confidence interval the model has for a respondent with no high school education in @fig-confimportances is significantly large, spanning from around -0.6 to +1.6, signifying an uncertainty among that demographic about who to vote for, those with higher education agree more on who to vote for. High school graduates slightly favour Biden, and those with University, College, or Post Graduate education are suggested by @tbl-modelsummaryimportances to be more likely to support Biden. As this relates to gun ownership as in @fig-gunownvseducation, we can also see here that the impact of owning a gun or not is offset by the respondents education. The votes are much closer between Biden and Trump for those with higher education and own guns compared to those without higher education and own guns.

@@

what is done and what we learned:::

In this paper we aimed to verify our expectations that ones cultural background and what one holds to personal importance affect their decision making process when it comes to elections. As expected, being of different cultural background impacted who they voted for. Exactly what those ideals that they chose to vote with we did not define here, but we know that correlation exists. Better defined here is the specific ideals that a respondent holds, modeling gun ownership and student loans and education levels in this paper. Also as expected we find that knowing a respondent's opinions on key political topics, in this case gun laws and student loan forgiveness, can allow us to predict who they are voting for, specifically here Trump who was in support of gun ownership laws and Biden who was in support of student loan forgiveness.

weakenesses of the analysis::: TODO also talk about small demographic sizes

@@

future steps to learn in the future:::

TODO talk about the change or no change that occurs in the 2024 election, and if respondents of different demographics had changed their minds after Biden's term. If people who had their student loans forgiven will still vote for Biden, or vote for Trump after their needs from Biden have already been met.

@@

```{r}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-modelsummaryrace
#| tbl-cap: "Explaining whether a voter supported Biden or Trump in 2020 given their race"

# Rename coefficients
cm <- c(
        'raceOther' = 'Other Race',
        'raceWhite' = 'White',
        'raceNative American' = 'Native American', 
        'raceMultiracial' = 'Multiracial',
        'raceMiddle Eastern' = 'Middle Eastern',
        'raceHispanic' = 'Hispanic',
        'raceBlack' = 'Black',
        # Since "Asian" comes first alphabetically, it was factored as '0', being therefore included in the intercept. 
        # as for this model only includes 
        '(Intercept)' = 'Asian'
        )

modelsummary(
  list(
    "Support Biden" = political_preferences_race
  ),
  statistic = "mad",
  coef_rename = cm
  )
```

@@

cm <- c('genderMale' = 'Male',
        'raceWhite' = 'White',
        'raceNative American' = 'Native American', 
        'raceMultiracial' = 'Multiracial',
        'raceMiddle Eastern' = 'Middle Eastern',
        'raceHispanic' = 'Hispanic',
        'raceBlack' = 'Black',
        'raceAsian' = 'Asian',
        'religionNone' = 'Non-Religious',
        'religionProtestant' = 'Protestant',
        'religionAgnostic' = 'Agnostic',
        'religionAtheist' = 'Atheist',
        'religionRoman Catholic' = 'Roman Catholic',
        'religionJewish' = 'Jewish',
        'religionOther' = 'Other',
        'age_group30-44' = 'Age (30-44)',
        'age_group45-65' = 'Age (45-64)',
        'age_group65+' = 'Age (65+)'
        )
